---
title: 'MatchIt: Getting Started'
author: "Noah Greifer"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: yes
    df_print: paged
  html_vignette:
    toc: yes
vignette: "%\\VignetteIndexEntry{MatchIt: Getting Started} \n%\\VignetteEngine{knitr::rmarkdown}
  \n%\\VignetteEncoding{UTF-8}\n"
bibliography: references.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      fig.width=7, fig.height=5)
options(width = 200)
```

```{=html}
<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>
```
## Introduction

`MatchIt` implements the suggestions of @ho2007 for improving parametric statistical models for estimating treatment effects in observational studies and reducing model dependence by preprocessing data with semi-parametric and non-parametric matching methods. After appropriately preprocessing with `MatchIt`, researchers can use whatever parametric model they would have used without `MatchIt` and produce inferences that are more robust and less sensitive to modeling assumptions. `MatchIt` reduces the dependence of causal inferences on commonly made, but hard-to-justify, statistical modeling assumptions using a large range of sophisticated matching methods. The package includes several popular approaches to matching and provides access to methods implemented in other packages through its single, unified, and easy-to-use interface.

Matching is used in the context of estimating the causal effect of a binary treatment or exposure on an outcome while controlling for measured pre-treatment variables, typically confounding variables or variables prognostic of the outcome. Here and throughout the `MatchIt` documentation we use the word "treatment" to refer to the focal causal variable of interest, with "treated" and "control" reflecting the names of the treatment groups. The goal of matching is to produce *covariate balance*, that is, for the distributions of covariates in the two groups to be approximately equal to each other, as they would be in a successful randomized experiment. The importance of covariate balance is that it allows for increased robustness to the choice of model used to estimate the treatment effect; in perfectly balanced samples, a simple difference in means can be a valid treatment effect estimate. Here we do not aim to provide a full introduction to matching or causal inference theory, but simply to explain how to use `MatchIt` to perform nonparametric preprocessing. For excellent and accessible introductions to matching, see @stuart2010 and Austin @austin2011b.

A matching analysis involves four primary steps: 1) planning, 2) matching, 3) assessing the quality of matches, and 4) estimating the treatment effect and its uncertainty. Here we briefly discuss these steps and how they can be implemented with `MatchIt`; in the other included vignettes, these steps are discussed in more detail.

We will use Lalonde's data on the evaluation of the National Supported Work program to demonstrate `MatchIt`'s capabilities. First, we load `MatchIt` and bring in the `lalonde` dataset.

```{r}
library("MatchIt")
data("lalonde")
head(lalonde)
```

The statistical quantity of interest is the causal effect of the treatment (`treat`) on 1978 earnings (`re78`). The other variables are pre-treatment covariates. See `?lalonde` for more information on this dataset. In particular, the analysis is concerned with the marginal, total effect of the treatment for those who actually received the treatment.

In what follows, we briefly describe the four steps of a matching analysis and how to implement them in `MatchIt`. For more details, we recommend reading the other vignettes, `vignette("matching-methods")`, `vignette("assessing-balance")`, and `vignette("estimating-effects")`, especially for users less familiar with matching methods. For the use of `MatchIt` with sampling weights, also see `vignette("sampling-weights")`. It is important to recognize that the ease of using `MatchIt` does not imply the simplicity of matching methods; advanced statistical methods like matching that require many decisions to be made and caution in their use should only be performed by those with statistical training.

## Planning

The planning phase of a matching analysis involves selecting the type of effect to be estimated, selecting the target population to which the treatment effect is to generalize, and selecting the covariates for which balance is required for an unbiased estimate of the treatment effect. Each of these are theoretical steps that do not involve performing analyses on the data. Ideally, they should be considered prior to data collection in the planning stage of a study. Thinking about them early can aid in performing a complete and cost-effective analysis.

**Selecting the type of effect to be estimated.** There are a few different types of effects to be estimated. In the presence of mediating variables, one might be interested in the direct effect of the treatment that does not pass through the mediating variables or the total effect of the treatment across all causal pathways. Matching is well suited for estimating total effects, and specific mediation methods may be better suited for other mediation-related quantities. One may be interested in a conditional effect or a marginal effect. A conditional effect is the effect of a treatment within some strata of other prognostic variables (e.g., at the patient level), and a marginal effect is the average effect of a treatment in a population (e.g., for implementing a broad policy change). Different types of matching are well suited for each of these, but the most common forms are best used for estimating marginal treatment effects; for conditional treatment effects, typically modeling assumptions are required or matching must be done within strata of the conditioning variables. Matching can reduce the reliance on correct model specification for conditional effects.

**Selecting a target population.** The target population is the population to which the effect estimate is to generalize. Typically, an effect estimated in a sample generalizes to the population from which the sample is a probability sample. If the sample is not a probability sample from any population (e.g., it is a convenience sample or involves patients from an arbitrary hospital), the target population can be unclear. Often, the target population is a group of units who are eligible for the treatment (or a subset thereof). Causal estimands are defined by the target population to which they generalize.

The average treatment effect in the population (ATE) is the average effect of the treatment for all units in the target population. The average treatment effect in the treated (ATT) is the average effect of the treatment for units like those who actually were treated. The most common forms of matching are best suited for estimating the ATT, though some are also available for estimating the ATE. Some matching methods distort the sample in such a way that the estimated treatment effect corresponds neither to the ATE nor to the ATT, but rather to the effect in an unspecified population (sometimes called the ATM, or average treatment effect in the remaining matched sample). When the target population is not so important (e.g., in the case of treatment effect discovery), such methods may be attractive; otherwise, care should be taken in ensuring the effect generalizes to the target population of interest. Different matching methods allow for different target populations, so it is important to choose a matching method that allows one to estimate the desired effect. See @greiferChoosingEstimandWhen2021 for guidance on making this choice.

**Selecting covariates to balance.** Selecting covariates carefully is critical for ensuring the resulting treatment effect estimate is free of confounding and can be validly interpreted as a causal effect. To estimate total causal effects, all covariates must be measured prior to treatment (or otherwise not be affected by the treatment). Covariates should be those that cause variation in the outcome and selection into treatment group; these are known as confounding variables. See @vanderweele2019 for a guide on covariate selection. Ideally these covariates are measured without error and are free of missingness.

## Check Initial Imbalance

After planning and prior to matching, it can be a good idea to view the initial imbalance in one's data that matching is attempting to eliminate. We can do this using the code below:

```{r}
# No matching; constructing a pre-match matchit object

```

The first argument is a `formula` relating the treatment to the covariates used in estimating the propensity score and for which balance is to be assessed. The `data` argument specifies the dataset where these variables exist. Typically, the `method` argument specifies the method of matching to be performed; here, we set it to `NULL` so we can assess balance prior to matching[^1]. The `distance` argument specifies the method for estimating the propensity score, a one-dimensional summary of all the included covariates, computed as the predicted probability of being the treated group given the covariates; here, we set it to `"glm"` for generalized linear model, which implements logistic regression by default[^2] (see `?distance` for other options).

[^1]: Note that the default for `method` is `"nearest"` to perform nearest neighbor matching. To prevent any matching from taking place in order to assess pre-matching imbalance, `method` must be set to `NULL`.

[^2]: Note that setting `distance = "logit"`, which was the default in `MatchIt` version prior to 4.0.0, will also estimate logistic regression propensity scores. Because it is the default, the `distance` argument can actually be omitted if logistic regression propensity scores are desired.

Below we assess balance on the unmatched data using `summary()`:

```{r}
# Checking balance prior to matching

```
